Appendix S1

### 2020-06-11 analysis/paper/appendix_s1.Rmd
minor text edits

To prevent overfitting, we pared down the more than 100 metrics generated by the CHaMP protocol describing the quantity and quality of fish habitat for each survey site. Habitat metrics were first grouped into broad categories that included channel unit, complexity, cover, riparian areas, side channels, size, substrate, temperature, water quality, and woody debris. Habitat metrics measuring any large wood volume were scaled by the site length (in 100 m units). To assist in determining the habitat metrics to include in the QRF model, we used the Maximal Information-Based Nonparametric Exploration (MINE) class of statistics [@Reshef2011] to determine those habitat characteristics (covariates) most highly associated with the log of observed parr densities. We calculated the maximal information coefficient (MIC), using the R package *minerva* [@R-minerva], to measure the strength of the linear or non-linear association between the natural log of fish density and each habitat metric [@Reshef2011]. MIC is a measure of correlation that incorporates potential non-linear associations, so for example if there is an expontential association the MIC value could be high, even when the standard correlation coefficient is low. We excluded categorical variables such as channel type (e.g. meandering, pool-riffle, plane-bed, etc.) because we assumed that other quantitative metrics would capture the differences between those qualitative categorical metrics.

Within each category, metrics were ranked according to their MIC value (Table \@ref(tab:mic-tab) and Figure \@ref(fig:mine-facet-fig)). The MIC value between each of the measured habitat characteristics and parr density was used to inform decisions on which habitat covariates to include in the QRF parr capacity model. We selected one or two variables amongst those with the highest MIC scores within each category, attempting to avoid covariates that were too highly correlated (Figure \@ref(fig:corr-fig)), while focusing on covariates we thought could influence fish behavior. For example, cumulative drainage area, mean annual flow and observed discharge are all highly correlated, but fish really only experience the observed flow, so we chose to include that metric in our QRF model. We also tried to include covariates that can be directly influenced by restoration actions or have been shown to impact salmonid juvenile density. Finally, we attempted to avoid metrics with too many missing values, or too many all zero values, in the data set, as well as metrics that may have too much observer error [@Rosgen2018].

@@

```{r mic-tab}
mine_tab_df = mine_res %>%
  filter(MetricCategory != 'Land Classification') %>%
  group_by(MetricCategory) %>%
  slice(1:10) %>%
  select(Category = MetricCategory,
         Name,
         Abbrv = Metric,
         MIC,
         perc_non0,
         selected)

mine_tab_df %>%
  select(-selected) %>%
  kable(digits = 3,
        format = 'html',
        caption = 'MIC statistic for top metrics in each habitat category. Metrics selected for the QRF model are in bold.') %>%
  kable_styling(fixed_thead = T) %>%
  row_spec(row = which(mine_tab_df$selected),
           bold = T)

@@

```{r mine-facet-fig, fig.cap = 'Barplots of MIC statistics, faceted by habitat category.'}
ggplot(data = mine_res,
       aes(x = Name,
           y = MIC)) +
  geom_col(fill = 'blue') +
  coord_flip() +
  facet_wrap(~ MetricCategory,
             scales = 'free_y',
             ncol = 3) +
  theme(legend.position = 'bottom',
        axis.text = element_text(size = 5))
```

>>>>

To prevent overfitting, we pared down the more than 100 metrics generated by the CHaMP protocol describing the quantity and quality of fish habitat for each survey site. Habitat metrics were first grouped into broad categories that included channel unit configuration, complexity, fish cover, riparian areas, side channels, stream size, substrate, temperature, water quality, and woody debris. Habitat metrics measuring any large wood volume were scaled by the site length (in 100 m units). To assist in determining the habitat metrics to include in the QRF model, we used the Maximal Information-Based Nonparametric Exploration (MINE) class of statistics [@Reshef2011] to determine those habitat characteristics (covariates) most highly associated with the log of observed parr densities. We calculated the maximal information coefficient (MIC), using the R package *minerva* [@R-minerva], to measure the strength of the linear or non-linear association between the natural log of fish density and each habitat metric [@Reshef2011]. MIC is a measure of correlation that incorporates potential non-linear associations, so for example if there is an quadratic association the MIC value could be high, even when the standard correlation coefficient is low. We excluded categorical variables such as channel type (e.g. meandering, pool-riffle, plane-bed, etc.) because we assumed that other quantitative metrics would capture the differences between those qualitative categorical metrics.

Within each category, metrics were ranked according to their MIC value (Table \@ref(tab:mic-tab) and Figure \@ref(fig:mine-facet-fig)). The MIC value of each measured habitat characteristics and parr density was used to inform decisions on which habitat covariates to include in the QRF parr capacity model. We selected one or two variables amongst those with the highest MIC scores within each category, attempting to avoid covariates that were too highly correlated (Figure \@ref(fig:corr-fig)), while focusing on covariates we thought could influence fish behavior. For example, cumulative drainage area, mean annual flow and observed discharge are all highly correlated, but fish really only experience the observed discharge, so we chose to include that metric in our QRF model. We also tried to include covariates that can be directly influenced by restoration actions or have been shown to impact salmonid juvenile density. Finally, we attempted to avoid metrics with too many missing values, or too many all zero values, in the data set, as well as metrics that may have too much observer error [@Rosgen2018].

@@

```{r mic-tab}
mine_tab_df = mine_res %>%
  mutate(`Percent 0-value` = 1 - perc_non0) %>%
  filter(MetricCategory != 'Land Classification') %>%
  group_by(MetricCategory) %>%
  slice(1:10) %>%
  select(Category = MetricCategory,
         Name,
         Abbrv = Metric,
         MIC,
         `Percent Missing` = perc_NA,
         `Percent 0-value`,
         selected)

mine_tab_df %>%
  select(-selected) %>%
  kable(digits = 3,
        format = 'html',
        # format = 'latex',
        booktabs = T,
        caption = 'MIC statistic for top metrics in each habitat category. Metrics selected for the QRF model are in bold.') %>%
  kable_styling(fixed_thead = T) %>%
  row_spec(row = which(mine_tab_df$selected),
           bold = T)

@@

```{r mine-facet-fig, fig.cap = 'Barplots of MIC statistics, faceted by habitat category.'}
mine_res %>%
  filter(MetricCategory != 'Land Classification') %>%
  mutate(MetricCategory = recode(MetricCategory,
                                 'ChannelUnit' = 'Channel\nUnit',
                                 'SideChannel' = 'Side\nChannel',
                                 'WaterQuality' = 'Water\nQuality')) %>%
  ggplot(aes(x = Name,
             y = MIC)) +
  geom_col(fill = 'blue') +
  coord_flip() +
  facet_wrap(~ MetricCategory,
             scales = 'free_y',
             ncol = 3) +
  theme(legend.position = 'bottom',
        axis.text = element_text(size = 5))
```
